---
title: "Course Project 1"
author: "Ankur Agarwal"
date: "September 9, 2015"
output: 
  html_document: 
    fig_width: 5
    toc: yes
---

#Course Project 1

This R Markdown document shows some important code snippets used in Exploratory Data Analysis: Course Project 1.

##Task 1: Reading Data

The original data is gigantic and we only need to use the data for dates: 1/2/2007 to 2/2/2007. There are lots of ways to do that. 

> Save the data once its read. Don't need to read it again and again.
Also make this a function which returns a data frame.

```{r, eval=FALSE}
#Method 1: Fast and too many steps
#We can grep for the required data
read.table(pipe('grep "^[1-2]/2/2007" "household_power_consumption.txt"'))
#This will need additional processing to split the single column into multiple columns but is pretty fast

#Method 2: Slow and too many steps (readDataFunc_lapply.R)
#We can read in only the date column 
colClasses = c('character', rep('NULL', 8))
#Then extract only the dates we require. Then read only the row indices by looping and skipping
requiredDates <- which(dataDates$Date %in% c("1/2/2007", "2/2/2007"))
d <- lapply(reqdDates, 
                function(x){read.table("household_power_consumption.txt", sep = ';',
                                       na.strings = '?',nrows = 1, header = F, as.is = T,
                                       skip = x, colClasses = classes, comment.char = "")})

#Method 3: Slow but no additional steps (readDataFunc_sqldf.R)
requiredData <- read.csv2.sql("household_power_consumption.txt", sql = "SELECT * FROM file
                              WHERE Date in ('1/2/2007', '2/2/2007')")
```
```{r message=FALSE, cache=TRUE}
source('readDataFunc_sqldf.R')
df <- readDataFunc()
```
```{r, eval=FALSE}
# The readDataFunc also pastes the date and time together and converts them from char to POSIX time format
requiredData$Date <- as.character.Date(requiredData$Date)
requiredData$Time <- as.character.Date(requiredData$Time)
requiredData$DateTime <- paste(requiredData$Date, requiredData$Time)
requiredData$DateTime <- strptime(requiredData$DateTime, "%d/%m/%Y %H:%M:%S")
```


##Task 2: Plotting

*   **Plot 1**  
Simple histogram
```{r}
hist(df$Global_active_power, col='red', 
     main = "Global Active Power", xlab = 'Global Active Power (kilowatts)')
```

*   **Plot 2**  
Line plot with Date on X axis
```{r}
plot(x = df$DateTime, y = df$Global_active_power, type='l',
     ylab = "Global Active Power (kilowatts)", xlab = "")

#these plots are automatically showing wdays in the x ticks, but if they don't we can change
#x tick positions and labels
```

*   **Plot 3**  
3 Line plots in 1
```{r}
plot(x = df$DateTime, y = df$Sub_metering_1, type='l',
     ylab = "Energy Sub Metering", xlab = "")
lines(x = df$DateTime, y = df$Sub_metering_2, col = 'red')
lines(x = df$DateTime, y = df$Sub_metering_3, col = 'blue')
legend("topright",lty = 1,
       col = c("black","blue", "red"), 
       legend = c("Sub_metering_1", "Sub_metering_2", "Sub_metering_3"))
```

*   **Plot 4**
4 subplots
```{r}
par(mfrow=c(2,2), mar = c(4,4,2,1), oma=c(0,0,0,0))
with(df, {
    #1,1
    plot(x = df$DateTime, y = df$Global_active_power, type='l',
         ylab = "Global Active Power", xlab = "");
    
    #2,1
    plot(x = df$DateTime, y = df$Voltage, type='l',
         ylab = "Voltage", xlab = "")
    title(sub = 'datetime', line = 3)
    
    #1,2
    plot(x = df$DateTime, y = df$Sub_metering_1, type='l',
         ylab = "Energy_Sub_Metering", xlab = "")
    lines(x = df$DateTime, y = df$Sub_metering_2, col = 'red')
    lines(x = df$DateTime, y = df$Sub_metering_3, col = 'blue');
    legend("topright",lty = 1,
           col = c("black","blue", "red"), 
           bty = 'n',
           cex = 0.8,
           legend = c("Sub_metering_1", "Sub_metering_2", "Sub_metering_3"))
    
    
    #2,2
    plot(x = df$DateTime, y = df$Global_reactive_power, type='l',
         ylab = "Global Reactive Power", xlab = "")
    title(sub = 'datetime', line = 3)
})
```

##Task 3: Saving to file
We need to open the right device to use. Remember to close once completed.
```{r eval=FALSE}
dev.copy(png, 'plot1.R', width = 640, height = 480)
dev.off()
```

