---
title: "R Basics"
author: "Ankur Agarwal"
date: "August 31, 2015"
output: 
  html_document: 
    fig_caption: yes
    toc: yes
---
This document is used to keep a record of basic (but important) things one comes across. Code snippets are also added for easy understanding and practise.

##Downloading/Reading/Writing a data frame
Some examples.
For certain file types the mode needs to be "wb"
```{r eval=FALSE}
download.file(fileUrl, destfile = fileName, mode="wb")
```
There are lots of functions for reading/ writing.


##Simulating a data frame

```{r}
d <- data.frame(id=sample(1:3, 10, T), val1 = rbinom(10, 10, 0.5), val2 = rnorm(10, mean = 5))

head(d,3)
```

data.frame 'd' has 3 columns: id, val1 and val2.

##Applying factors from another data frame
Now say, the 'id' column in previous data set refers to some names; stored in another dataframe.
```{r}
dname <- data.frame(id = 1:3,name = c("A","B", "C"))

dname
```
We can replace the values in d\$id with the values in dname\$name.
First, we'll need to factor both, so that they can be replaced w/o for loops.

```{r}
d$id <- factor(d$id, 1:3)
dname$name <- factor(dname$name)
levels(d$id) <- levels(dname$name)
head(d,3)
```

Now we can tabulate this data frame with the names instead of numbers, making it for easy reading.
```{r}
table(d$id)
table(d$id, d$val1)
```

#JSON
Usefule when receiving JSON based replies.
```{r eval=FALSE}
#Receive the GET request in 'req'
data <- content(req) 
#The data can be converted to JSON form
JSONData <- toJSON(data)
#This is data in JSON form 
#If the data has a good structure, convert to a data.frame
df <- fromJSON(JSONData)
```

#Finding/ counting TRUE indices
```{r}
d <- sample(LETTERS[1:3], 10, replace = T)
print(d)
#sum will return total number of occurences that satisfy a condition
sum(d == 'A')
#which will return the actual indices
which(d == 'A')
```

#String Operations
These are just representatives. Look in 'help' or online.
```{r}
#strsplit: split elements of a character vector
#Example taken from help ?strsplit
#uses regex. so '.' means any character.
strsplit("a.b.c", ".")
#Needs some special treatment 
strsplit("a.b.c", "[.]")
strsplit("a.b.c", ".", fixed = TRUE)

#substitute
#Also a regexp (regex expression)
gsub("[.]", " ", "gmail.com")

#grep
#self explanatory
grep("[jJ]une", c("June","July", "August"), value = T)
```



