---
title: "R Basics"
author: "Ankur Agarwal"
date: "August 31, 2015"
output: 
  html_document: 
    fig_caption: yes
    toc: yes
---
This document is used to keep a record of basic (but important) things one comes across. Code snippets are also added for easy understanding and practise.

##Simulating a data frame

```{r}
d <- data.frame(id=sample(1:3, 10, T), val1 = rbinom(10, 10, 0.5), val2 = rnorm(10, mean = 5))

head(d,3)
```

data.frame 'd' has 3 columns: id, val1 and val2.

##Applying factors from another data frame
Now say, the 'id' column in previous data set refers to some names; stored in another dataframe.
```{r}
dname <- data.frame(id = 1:3,name = c("A","B", "C"))

dname
```
We can replace the values in d\$id with the values in dname\$name.
First, we'll need to factor both, so that they can be replaced w/o for loops.

```{r}
d$id <- factor(d$id, 1:3)
dname$name <- factor(dname$name)
levels(d$id) <- levels(dname$name)
head(d,3)
```

Now we can tabulate this data frame with the names instead of numbers, making it for easy reading.
```{r}
table(d$id)
table(d$id, d$val1)
```

